### 2022-07-14 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@793e17ad9ae1f8e38ec8aea9df678034c42dd448 🚀

In this contribution, we used two complementary approaches to predict
the structure of food webs using the principle of maximum entropy. We
then compared our predictions against empirical data and null and
neutral models commonly used in network ecology. The first approach
consists in deriving constrained probability distributions of given
network properties directly. We derived the joint degree distribution (a
probability distribution) of maximum entropy using only the number of
species \(S\) and the number of interactions \(L\) as state variables.
Then, we predicted the degree distribution of maximum entropy directly
from the joint degree distribution since the first is the sum of the
marginal distributions of the second (a species' degree is the sum of
its in and out-degrees). Because of the scarcity of empirical data on
the number of links in ecological networks, in Box 2 we present a method
to predict \(L\) from \(S\), thus allowing the prediction of the joint
degree distribution from \(S\) solely. In turn, the second approach
consists in finding, using different constraints, the adjacency matrix
of maximum entropy from which network properties can be measured. To do
so, we used a flexible and heuristic approach based on simulating
annealing to find networks \emph{close} to maximum entropy. As discussed
above, our choice of algorithm stands from the very nature of food webs
(i.e., simple directed networks allowing self-loops) that makes the
analytical derivation of a maximum entropy graph model difficult. We
first built our type I MaxEnt network model constrained by the
connectance of the network (i.e., the ratio \(L/S^2\)). A comparison of
this model against empirical data indicated that connectance alone was
not sufficient to predict many aspects of network structure. For this
reason, we built our type II MaxEnt network model, which instead uses
the whole joint degree sequence as a constraint. Overall, we found that
this second model was much better at predicting food-web structure than
the one based on connectance.

@@

We tested our MaxEnt models (both approaches) against open food-web data
queried from three different sources and integrated into what we call
our \emph{complete dataset}. First, all food webs archived on
\texttt{mangal.io} were directly queried from the database
(\(n = 235\)). Most ecological networks archived on Mangal are
multilayer networks, i.e.~networks that describe different types of
interactions. We kept all networks whose interactions were mainly of
predation and herbivory types, and removed the largest network
(\(S = 714\)) for computational efficiency reasons. Then, to this set we
added food webs from two different sources: the New-Zealand dataset
(\(n = 21\); Pomeranz et al. 2018) and the Tuesday lake dataset
(\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Of these two datasets,
\(19\) networks had data on species' relative abundances that were used
in the neutral model presented in a following subsection. These networks
are part of what we call our \emph{abundance dataset}, which is a subset
of our complete dataset.

@@

We built two types of MaxEnt network models: one based on connectance
(type I MaxEnt network model) and the other based on the joint degree
sequence (type II MaxEnt network model). They are based on the same
constraints as the types I and II null models presented above. For both
models, we used a simulating annealing algorithm with \(4\) chains,
\(2000\) steps and an initial temperature of \(0.2\). For each chain, we
first generated one random Boolean matrix with the same order (number of
species) as empirical webs, while maintaining the total number of
interactions (type I MaxEnt network model) or rows and columns sums
(type II MaxEnt network model). These are our initial configurations.
Then, we swapped interactions sequentially while maintaining the
original connectance or the joint degree sequence for types I and II
MaxEnt network model, respectively. Configurations with a higher SVD
entropy than the previous one in the chain were always accepted, whereas
they were accepted with a probability conditional to a decreasing
temperature when lower. The final configuration with the highest SVD
entropy among the four chains constitute our estimated MaxEnt network.
Even though we decided to work with point estimates, it is possible to
have a (non MaxEnt) probability distribution of networks when working
with the entire chains after burn-in. For each network in our complete
and abundance datasets, we estimated their configuration with maximum
entropy using both types of MaxEnt network models.

@@

\begin{figure}
\hypertarget{fig:motifs}{%
\centering
\includegraphics{figures/motifs_distribution.png}
\caption{Motifs profile of empirical, maximum entropy, and null food
webs. Empirical networks include all food webs archived on Mangal, as
well as the New-Zealand and Tuesday lake datasets (our complete
dataset). Maximum entropy networks were estimated using a simulating
annealing algorithm to find the network of maximum SVD entropy while
maintaining the connectance (type I MaxEnt network model) or the joint
degree sequence (type II MaxEnt network model). The predictions of the
types I and II null models are also plotted. Boxplots display the median
proportions of each motif (middle horizontal lines), as well as the
first (bottom horizontal lines) and third (top horizontal lines)
quartiles. Vertical lines encompass all data points that fall within
\(1.5\) times the interquartile range from both quartiles, and dots are
data points that fall outside this range. Motifs names are from Stouffer
et al. (2007).}\label{fig:motifs}
}
\end{figure}

\begin{figure}
\hypertarget{fig:motifs_rel}{%
\centering
\includegraphics{figures/motifs_relations.png}
\caption{Relationship between motifs proportions of empirical, maximum
entropy, and null food webs. Empirical networks include all food webs
archived on Mangal, as well as the New-Zealand and Tuesday lake datasets
(our complete dataset). Maximum entropy networks were estimated using a
simulating annealing algorithm to find the network of maximum SVD
entropy while maintaining the connectance (type I MaxEnt network model)
or the joint degree sequence (type II MaxEnt network model). The
predictions of the types I and II null models are also plotted.
Regression lines are plotted in each panel. Motifs names are from
Stouffer et al. (2007).}\label{fig:motifs_rel}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:measures}{%
\centering
\includegraphics{figures/measures_emp_maxent.png}
\caption{Relationship between the structure of empirical and maximum
entropy food webs. Empirical networks include all food webs archived on
Mangal, as well as the New-Zealand and Tuesday lake datasets (our
complete dataset). Maximum entropy networks were estimated using a
simulating annealing algorithm to find the network of maximum SVD
entropy while maintaining the joint degree sequence (type II MaxEnt
network model). (a) Nestedness (estimated using the spectral radius of
the adjacency matrix), (b) the maximum trophic level, (c) the network
diameter, and (d) the SVD entropy were measured on these empirical and
maximum entropy food webs. The identity line is plotted in each
panel.}\label{fig:measures}
}
\end{figure}

@@

@@

>>>>

In this contribution, we used two complementary approaches to predict
the structure of food webs using the principle of maximum entropy. We
then compared our predictions against empirical data and null and
neutral models commonly used in network ecology. The first approach
consists in deriving constrained probability distributions of given
network properties directly. We derived the joint degree distribution (a
probability distribution) of maximum entropy using only the number of
species \(S\) and the number of interactions \(L\) as state variables.
Then, we predicted the degree distribution of maximum entropy directly
from the joint degree distribution since the first is the sum of the
marginal distributions of the second (a species' degree is the sum of
its in and out-degrees). Because of the scarcity of empirical data on
the number of links in ecological networks, in Box 2 we present a method
to predict \(L\) from \(S\), thus allowing the prediction of the joint
degree distribution from \(S\) solely. In turn, the second approach
consists in finding, using different constraints, the adjacency matrix
of maximum entropy from which network properties can be measured. To do
so, we used a flexible and heuristic approach based on simulated
annealing to find networks \emph{close} to maximum entropy. As discussed
above, our choice of algorithm stands from the very nature of food webs
(i.e., simple directed networks allowing self-loops) that makes the
analytical derivation of a maximum entropy graph model difficult. We
first built our type I MaxEnt network model constrained by the
connectance of the network (i.e., the ratio \(L/S^2\)). A comparison of
this model against empirical data indicated that connectance alone was
not sufficient to predict many aspects of network structure. For this
reason, we built our type II MaxEnt network model, which instead uses
the whole joint degree sequence as a constraint. Overall, we found that
this second model was much better at predicting food-web structure than
the one based on connectance.

@@

We tested our MaxEnt models (both approaches) against open food-web data
queried from three different sources and integrated into what we call
our \emph{complete dataset}. First, all food webs archived on
\texttt{mangal.io} (Poisot et al. 2016; Banville, Vissault, and Poisot
2021) were directly queried from the database (\(n = 235\)). Most
ecological networks archived on Mangal are multilayer networks,
i.e.~networks that describe different types of interactions. We kept all
networks whose interactions were mainly of predation and herbivory
types, and removed the largest network (\(S = 714\)) for computational
efficiency reasons. Then, to this set we added food webs from two
different sources: the New-Zealand dataset (\(n = 21\); Pomeranz et al.
2018) and the Tuesday lake dataset (\(n = 2\); Cohen, Jonsson, and
Carpenter 2003). Of these two datasets, \(19\) networks had data on
species' relative abundances that were used in the neutral model
presented in a following subsection. These networks are part of what we
call our \emph{abundance dataset}, which is a subset of our complete
dataset.

@@

We built two types of MaxEnt network models: one based on connectance
(type I MaxEnt network model) and the other based on the joint degree
sequence (type II MaxEnt network model). They are based on the same
constraints as the types I and II null models presented above. For both
models, we used a simulated annealing algorithm with \(4\) chains,
\(2000\) steps and an initial temperature of \(0.2\). For each chain, we
first generated one random Boolean matrix with the same order (number of
species) as empirical webs, while maintaining the total number of
interactions (type I MaxEnt network model) or rows and columns sums
(type II MaxEnt network model). These are our initial configurations.
Then, we swapped interactions sequentially while maintaining the
original connectance or the joint degree sequence for types I and II
MaxEnt network model, respectively. Configurations with a higher SVD
entropy than the previous one in the chain were always accepted, whereas
they were accepted with a probability conditional to a decreasing
temperature when lower. The final configuration with the highest SVD
entropy among the four chains constitute our estimated MaxEnt network.
Even though we decided to work with point estimates, it is possible to
have a (non MaxEnt) probability distribution of networks when working
with the entire chains after burn-in. For each network in our complete
and abundance datasets, we estimated their configuration with maximum
entropy using both types of MaxEnt network models.

@@

\begin{figure}
\hypertarget{fig:motifs}{%
\centering
\includegraphics{figures/motifs_distribution.png}
\caption{Motifs profile of empirical, maximum entropy, and null food
webs. Empirical networks include all food webs archived on Mangal, as
well as the New-Zealand and Tuesday lake datasets (our complete
dataset). Maximum entropy networks were estimated using a simulated
annealing algorithm to find the network of maximum SVD entropy while
maintaining the connectance (type I MaxEnt network model) or the joint
degree sequence (type II MaxEnt network model). The predictions of the
types I and II null models are also plotted. Boxplots display the median
proportions of each motif (middle horizontal lines), as well as the
first (bottom horizontal lines) and third (top horizontal lines)
quartiles. Vertical lines encompass all data points that fall within
\(1.5\) times the interquartile range from both quartiles, and dots are
data points that fall outside this range. Motifs names are from Stouffer
et al. (2007).}\label{fig:motifs}
}
\end{figure}

\begin{figure}
\hypertarget{fig:motifs_rel}{%
\centering
\includegraphics{figures/motifs_relations.png}
\caption{Relationship between motifs proportions of empirical, maximum
entropy, and null food webs. Empirical networks include all food webs
archived on Mangal, as well as the New-Zealand and Tuesday lake datasets
(our complete dataset). Maximum entropy networks were estimated using a
simulated annealing algorithm to find the network of maximum SVD entropy
while maintaining the connectance (type I MaxEnt network model) or the
joint degree sequence (type II MaxEnt network model). The predictions of
the types I and II null models are also plotted. Regression lines are
plotted in each panel. Motifs names are from Stouffer et al.
(2007).}\label{fig:motifs_rel}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:measures}{%
\centering
\includegraphics{figures/measures_emp_maxent.png}
\caption{Relationship between the structure of empirical and maximum
entropy food webs. Empirical networks include all food webs archived on
Mangal, as well as the New-Zealand and Tuesday lake datasets (our
complete dataset). Maximum entropy networks were estimated using a
simulated annealing algorithm to find the network of maximum SVD entropy
while maintaining the joint degree sequence (type II MaxEnt network
model). (a) Nestedness (estimated using the spectral radius of the
adjacency matrix), (b) the maximum trophic level, (c) the network
diameter, and (d) the SVD entropy were measured on these empirical and
maximum entropy food webs. The identity line is plotted in each
panel.}\label{fig:measures}
}
\end{figure}

@@

\leavevmode\hypertarget{ref-Banville2021ManJla}{}%
Banville, Francis, Steve Vissault, and Timothée Poisot. 2021.
{``Mangal.jl and EcologicalNetworks.jl: Two Complementary Packages for
Analyzing Ecological Networks in Julia.''} \emph{Journal of Open Source
Software} 6 (61): 2721. \url{https://doi.org/10.21105/joss.02721}.

@@

\leavevmode\hypertarget{ref-Poisot2016ManMakb}{}%
Poisot, Timothee, Benjamin Baiser, Jennifer A. Dunne, Sonia Kefi,
Francois Massol, Nicolas Mouquet, Tamara N. Romanuk, Daniel B. Stouffer,
Spencer A. Wood, and Dominique Gravel. 2016. {``Mangal - Making
Ecological Network Analysis Simple.''} \emph{Ecography} 39 (4): 384--90.
\url{https://doi.org/10.1111/ecog.00976}.
